---
title: "R Notebook"
output: html_notebook
---

In this File, we are dealing with the *Singleton* Time points and the Reporting of Matching and Non-Matching Subject IDs from the *BMI* and *Media* data sets.

#### R Libraries

This code block has all the needed R libraries

```{r}
#For the dta raw files
library(foreign)
#For importing different types of data set without specification
library(rio)
#For processing long form data
library(dplyr)
#GTools library for ordering numeric variables
library(gtools)
```


#### Uploading Raw Singleton data

In this code chunk, we are uploading the excluded singleton data from BMI and Media to see if there are any matches between them.

```{r}
bmi_singleton <- import("../../data/final/bmisubjects_withonedatap.csv")
media_singleton <- import("../../data/final/mediasubjects_withonedatap.csv")
```

#### Matching the Singletons

```{r}
#Checking how many command IDs are between Singleton BMI and Singleton Media

common_ID <- intersect(bmi_singleton$ID_, media_singleton$ID_)
print(common_ID)

#A single singleton subject ID of 626 matches!
```

#### Uploading Raw Data to explore Subject ID matches and mismatches

```{r}


```






