---
title: "02_InitialModels"
author: "Zarni Htet"
date: "March 7, 2018"
output: github_document
---

### Libraries for the lm and lmer packages
```{r}
library(lme4) # For lmer function
library(lmerTest)
library(rio) # For importing data frames
```

### Importing the data frames to be used
```{r}
bmi_media <- import("../../data/final/final_interp_data.csv")
bmi_media <- bmi_media[,-1]
```



### Functional Form

#### Simple Linear Regression

We are regressing BMI on Media and Time.

$$
Y(BMI) \sim X (Media)+ t (Months) + \epsilon   
$$

```{r}
fit <- lm(zBMI ~ Media + Months, data = bmi_media)
print(summary(fit))
```

#### Simple Linear Regression Interpretation

On average, a unit increase in media is associated with a 0.02868 increase in BMI. 

#### Multi Level Regression 

##### Varying Intercept Model with individual predictor

$$
BMI_{ti} = b_{0} + b_{1}Media_{ti} + b_{2}Months_{ti} + \zeta_{i} + \epsilon_{ti}
$$

*Note to self: Need to write the distribution of Zeta and its error distribution*

Lmer Code
```{r}
M1 = lmer(zBMI~Media + Months + (1|ID), data = bmi_media)
print(summary(M1))
```

#### Model Interpretation

Controlling for differences between subjects, the effect of one unit change in Media is 0.0161 to BMI and it is not statistically significant assuming our model assumptions are correct. With the likelihood ratio test below, the random effects are warranted.

#### Likelihood ratio test

```{r}
rand(M1)
```







