---
title: "RCT Outputs"
output: html_notebook
---

1) 

**Problem Set**

Show a treatment effect of 5 unit increase in Causal Inference Quiz I Score take II after the intervention of a home tutor program that is randomly assigned to students taking Professor Hill's class in Fall of 2018 .The home tutors (treatments) are equally talented in **All Ways** of graduate students by the name of Boss Andrea, Raven Ray and Zany Htet.  

**Assumptions**
Class size N is 100
The pre-treatment Causal Quiz Score is N(80, 10)
Treatment assigned is completely random.

**Conditions on Quiz 1 Take 2**
Both treated and non-treated score improves! 

**Expected Output**
A Linear Regression model for Causal Inference. Coefficient may round.

**Hint**
Build a simulated data set first

#### Simulated Data Set

Creating the Vectors for the Data Set
```{r}
#Create 100 students
students <- c(1:100)
#Create pre-treatment test scores for everyone
pretest <- rnorm(n=100, mean = 80, sd = 10)
#Create a vector of 0 and 1
zero <- rep(0,50)
one  <- rep(1,50)
zone <- c(zero, one)
#Creat a Randomly Assigned Treatment Variable
treat <- sample(zone, 100, replace = FALSE)
#Create a non-treated test score post-intervention
postest_nt <- rnorm(n=50, mean = 82, sd = 10)
#Create a treated test score post-intervention
postest_t <- rnorm(n =50, mean = 87, sd = 10)
```

#### Gluing the Data Together

```{r}



```

### Running the LM model








